### 2016-06-09 GPCE2016/sections/Appendix_Translation.tex
preparing for GPCE

\section{What  \mixin Generates}\label{sec:translation}

This section shows what the \mixin annotation generates and presents a
formal definition for most of the generated methods. Since the formalized part
of Classless Java does not consider
% setters, so we do not formally define the generation of the setters.
%We do not include
casts or \Q@instanceof@, the \Q@with@ method is not included in the
formal translation. For the same reason \Q@void@ returning setters are
not included, since they are just a minor variation over the more
interesting fluent setters, and they would require special handling
just for the conventional \Q@void@ type.

\subsection{Translation}
For the purposes of the formalization, the translation is divided into
two parts for more convenient discussion on formal properties later. To this aim we introduce the annotation
$\weakAnn$. Its role is only in the translation process, hence is
not part of the Classless Java language.  $\weakAnn$ generates the
constructor method $\QM{of}$, while \mixin automatically refines the
return types and calls $\weakAnn$.

\noindent$\begin{array}{l}
\InTextDef{16em}{[\![\mixinAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC ]\!]
}{
[\![\weakAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC
\methods\ \methods' \cC
]\!]}\\
\InTextWith{\methods'=\otherMethod(\C_0,\methods)}\\

\InTextDef{16em}{[\![\weakAnn\ \QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC \methods\ \cC ]\!]
}{
\QM{interface}\ \C_0\ \QM{extends}\ \Cs\ \oC
\methods\ \ofMethod(\C_0) \cC
}\\
\InTextWith{\valid(\C_0),\QM{of}\notin\dom(\C_0) }
\end{array}$

\noindent Note that it is necessary to explicitly check if the interface is valid
for annotation:

\noindent$\begin{array}{ll}
\InTextDef{7ex}{\valid(\C_0)}{\forall \m\in\dom(\C_0),\mbox{ if }\mh\QM; =
            \mBody(\m, \C_0),\mbox{ one of the following cases is satisfied:}}\\
\tab\tab
\isField(\method), \isWith(\method, \C_0) \mbox{ or }
\isSetter(\method,\C_0)\\
\InTextDef{24ex}{\isField(\C\ \m\oR\cR\QM;)}{
\mnot\ \specialName(\m)}\\
\InTextDef{24ex}{\isWith(\C'\ \QM{with#}\m \oR \C\ \x\cR\QM;, \C_0)}{
\C_0 <: \C', \mBody(\m, \C_0) = \C\ \m\oR\cR\QM;
\ \mand\ \mnot\ \specialName(\m)}\\
\InTextDef{24ex}{\isSetter(\C'\ \QM_\m \oR \C\ \x\cR\QM;, \C_0)}{
\C_0 <: \C', \mBody(\m, \C_0) = \C\ \m\oR\cR\QM;
\ \mand\ \mnot\ \specialName(\m)}\\

@@

\noindent$\begin{array}{ll}
\InTextDef{16em}{\C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\QM;\in
\otherMethod(\C_0,\methods)}{
 \isWith(\mBody(\QM{with#}\m, \C_0), \C_0)}\\
\hspace{2.65in}{
\mbox{ and } \QM{with#}\m\notin\dom(\methods)}\\
\InTextDef{16em}{\C_0\ \QM_\m\oR \C\ \QM{_val}\cR\QM;\in
\otherMethod(\C_0,\methods)}{\isSetter(\mBody(\QM_\m, \C_0), \C_0),}\\
\hspace{2.65in}{
\mbox{ and } \QM_\m\notin\dom(\methods)}\\
\end{array}$

@@

\paragraph{The $\ofMethod$ function:}\label{subsec:ofmethod}
The function $\ofMethod$ generates the method \QM{of}, as an object factory. To
avoid boring digressions into well-known ways to find unique
names, %for the sake of this formalization
we assume that all methods with no parameters do not start with an underscore,
and we prefix method names with underscores to obtain valid parameter names for
\QM{of}.

\noindent$\begin{array}{l}
\InTextDef{5em}{\ofMethod(\C_0)}{
 \QM{static}\ \C_0\ \QM{of} \oR \C_1\ \QM_\m_1\QM,\ldots \C_n\ \QM_\m_n\cR\
\QM{\{}
\QM{return new}\ \C_0 \oR\cR\ \QM{\{} }\\
\tab\tab \C_1\ \m_1 = \QM_\m_1\QM;\ldots \C_n\ \m_n = \QM_\m_n\QM; \\
\tab\tab
\C_1\ \m_1\oR\cR\ \QM{\{return }\ \m_1\QM{;\}}\ \ldots
\C_n\ \m_n\oR\cR\ \QM{\{return }\ \m_n\QM{;\}}\\
\tab\tab\withMethod(\C_1,\m_1,\C_0,\es_1)\ldots\withMethod(\C_n,\m_n,\C_0,\es_n)\\
\tab\tab\setterMethod(\C_1,\m_1,\C_0)\ldots\setterMethod(\C_n,\m_n,\C_0)\\
%\tab\tab\cloneMethod(\C_0,\es)\\
%\tab\tab\withMethod(\C_0)\\
\tab\QM{\};\}} \\
\InTextWith{\C_1\ \m_1\QM{();},\ldots \C_n\ \m_n\QM{();} = \fieldsFunc(\C_0) 
\mbox{ and }\es_i=\m_1\QM,\ldots\QM, \m_{i-1}\QM,\QM{_val,}\m_{i+1}\QM,\ldots\QM, \m_n}
\end{array}$

@@

\noindent$\begin{array}{ll}
\InTextDef{24ex}{\method\in\fieldsFunc(\C_0)}{
\isField(\method)\ \mand\
%}\\\InTextWith{
\method=\mBody(m^\method,\C_0)
}
\end{array}$

@@

\noindent$\begin{array}{l}
\InTextDef{10em}{\withMethod(\C,\m,\C_0,\es)}{
\C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\ \QM{\{}
\QM{return}\ \C_0\QM{.of(}\es\QM{);\}}} \\
\InTextWith{\mBody(\QM{with#}\m,\C_0) \mbox{ having the form }\mh\QM;}\\
\InTextDef{10em}{\withMethod(\C,\m,\C_0,\es)}{\emptyset\mbox{ otherwise}}\\
\InTextDef{10em}{\setterMethod(\C,\m,\C_0)}{
\C_0\ \QM_\m\oR \C\ \QM{_val}\cR\ \QM{\{}
 \m\QM{= _val;return this;\}}} \\
\InTextWith{
\mBody(\QM_\m,\C_0) \mbox{ having the form }\mh\QM;}\\
\InTextDef{10em}{\setterMethod(\C,\m,\C_0)}{\emptyset\mbox{ otherwise}}\\
%\cloneMethod:&\cloneMethod(\C_0,\es)=
%\C_0\ \QM{clone()\{return}\ \C_0\QM{.of(}\es\QM{);\}} \\
%&\mbox{iff }
%\mBody(\QM{clone},\C_0) \mbox{ is of form }\mh\QM;\\
%&\cloneMethod(\C_0,\es)=\emptyset\mbox{ otherwise}\\
\end{array}$

\subsection{Other Features}

@@

\subsection{Guarantees}
To understand to what extent our approach is correct (and more in general, what it means to say that a language tuning is correct) we identify three types of guarantees:
\begin{itemize}
\item \textit{Self coherence}: the generated code itself is well-typed; that is,
  type errors are not present in code the user has not written.
In our case it means that either \mixin{} produces (in a controlled way) an
understandable error, or the interface can be successfully annotated and the generated code is well-typed.
We guarantee \textit{Self coherence}.

\item \textit{Client coherence}: all the client code (for example method calls)
  that is well-typed before code generation is also well-typed after the generation.
The annotation just adds more behavior without removing any functionality.
We guarantee \textit{Client coherence}.

\item \textit{Heir coherence}: interfaces (and in general classes) inheriting the instrumented code are well-typed if they were well-typed without the instrumentation.
This would forbid adding any (default or abstract) method to the annotated interfaces, including type refinement.
\mixin  does not guarantee \textit{Heir coherence}.
Indeed consider the following:

\begin{lstlisting}
interface A { int x(); A withX(int x); }
@Obj interface B extends A {}
interface C extends B { A withX(int x); }
\end{lstlisting}

\noindent This code is correct before the translation, but \mixin would  generate in \Q@B@  a method ``\Q@B withX(int x);@''.
This would break \Q@C@. \\*
Similarly, an expression of the form ``\Q@new B(){.. A withX(int x){..}}@''
would be correct before translation, but would be ill-typed after the translation.
\end{itemize}

\noindent This means that our automatic type refinement
is a useful and convenient feature, but not transparent to the heirs of the annotated interface. They need to be aware of the annotation semantics and provide the right type while refining methods.

\subsection{Results}
To formally characterize the behavior of our annotation and the two levels of guarantees that we offer, we provide some notations and two theorems:
\begin{itemize}
\item We denote with $\C^{\II}$ the name of an interface.
\item An interface table
IT is OK if under such interface table, all interfaces are OK.
\item Since interface tables are just represented as sequences of interfaces we write IT = $\II$ IT' to select a specific interface in a table.
\item IT contains an heir of $\C$ if there is an interface that extends it, or a \Q@new@ that instantiate it.
\end{itemize}

\begin{thm}[@ObjOf tuning]
If a given interface table $\II$ IT is OK
 where $\II$ has $\weakAnn$,
$\valid(\C^{\II})$  and $\QM{of}\notin\dom(\C^{\II})$,
then the interface table $[\![\II]\!]$ IT is OK.
\end{thm}

\begin{thm}[@Obj tuning]
If a given interface table $\II$ IT is OK
 where $\II$ has $\mixinAnn$,
$\valid(\C^{\II})$  and $\QM{of}\notin\dom(\C^{\II})$, and there is no heir of $\C^{\II}$,
then the interface table $[\![\II]\!]$ IT is OK.
\end{thm}

Informally, the theorems mean that for a client program that
typechecks before the translation is applied, if the annotated type has
no subtypes and no objects of that type are created, then type safety
is guaranteed after the successful translation.

The second
step of \mixin, namely what $\weakAnn$ does in the formalization, is
guaranteed to be type-safe for the three kinds of coherence by the $\weakAnn$  tuning theorem.
\mixin
tuning is more interesting: since \mixin does not guarantee heir
coherence, we explicitly exclude the presence of heirs. In this way
\mixin tuning guarantees only self and client coherence. The formal
theorem proofs are available in the supplementary materials.%~\ref{subsec:lemma1},~\ref{subsec:lemma2} and~\ref{subsec:theorem}.\\

\begin{comment}
\begin{figure}[tbp]
\centering
\includegraphics[width=5in]{screenshot.png}
\caption{Screenshot.}\label{screenshot_png}
\end{figure}

\haoyuan{I tried to understand the current algorithm, and did more experiments in eclipse.
Now I borrow some ideas from the current version, and give a new version of the algorithm in text. See below.

(1) I guess the function \textsf{tops} is not necessary. The first step is still
\[\textsf{mbody}(m,C_i)\in\overline{meth}\textrm{ (excluding \textbf{static} methods)}\]

(2) Assume the context is ``interface $C_0$ extends $\overline{C}$ \{$meth'$;...\}''. First handle
\[\textsf{override}(meth',\overline{meth}) \eqno{(*)}\]

(3) If $meth'\ne\none$, $(*)$ returns $meth'$ if
\[\forall meth\in\overline{meth},meth'\subtype meth\]
even if there are conflicts in $\overline{meth}$.

(4) If $meth'=\none$, we need to figure out
\[\textsf{mostSpecific}(\overline{meth})\]
and it should be the one that ``overrides'' all the others in $\overline{meth}$. It means we should not only deal with the return types of methods, but also look into the subtyping relation of interfaces. But for abstract methods, only return types are taken into consideration.
}
\end{comment}

%\text{\yanlin{shouldn't mostSpecific be: $\forall \method' \in \methods : \method \subtype
%  \method'$ ?}}

%(2) If $body_1.\textsf{returnType}=body_2.\textsf{returnType}$, \textsf{shadow} tends to return a default method. If both $body_1$ and $body_2$ are default methods, \textsf{shadow} throws an error.
%\begin{equation*}
%\begin{array}{ll}
%\textsf{shadow}(body_1, body_2)=\textsf{ERROR} & \textsf{if }body_1.\textsf{modifier}=body_2.\textsf{modifier}=\textbf{default}\\
%\textsf{shadow}(body_1, body_2)=body_1 \hspace{.1in} & \textsf{if }body_1.\textsf{modifier}=\textbf{default} \\
%\textsf{shadow}(body_1, body_2)=body_2 \hspace{.1in} & \textsf{if }body_2.\textsf{modifier}=\textbf{default} \\
%\textsf{shadow}(body_1, body_2)=body_1\textsf{ (or }body_2\textsf{)} \hspace{.1in} & \textsf{otherwise}
%\end{array}
%\end{equation*}
%
%(3) If $body_1.\textsf{returnType}<:body_2.\textsf{returnType}$, \textsf{shadow} tends to choose the one with the subtype (namely $body_1$), but only when both methods are abstract, otherwise it gives an error. The other direction $body_2.\textsf{returnType}<:body_1.\textsf{returnType}$ follows the same rule. It also gives an error if there is no subtyping relationship between two return types.
%\begin{equation*}
%\begin{array}{ll}
%\textsf{shadow}(body_1, body_2)=body_1 & \textsf{if }body_1.\textsf{modifier}=body_2.\textsf{modifier}=\emptyset\\
%& \textsf{and }body_1.\textsf{returnType}<:body_2.\textsf{returnType}\\
%\textsf{shadow}(body_1, body_2)=body_2 & \textsf{if }body_1.\textsf{modifier}=body_2.\textsf{modifier}=\emptyset\\
%& \textsf{and }body_2.\textsf{returnType}<:body_1.\textsf{returnType}\\
%\textsf{shadow}(body_1, body_2)=\textsf{ERROR} \hspace{.1in} & \textsf{otherwise}
%\end{array}
%\end{equation*}

%\subsubsection{Auxiliary function: \textsf{replace}}
%
%The \textsf{replace} function takes two same methods (with the same name and types of arguments), and gives the result of the first method overriding the second one.
%
%\begin{equation*}
%\begin{array}{ll}
%\textsf{replace}(body_1, body_2)=body_1 & \textsf{if }body_2=\emptyset\\
%\textsf{replace}(body_1, body_2)=body_2 & \textsf{if }body_1=\emptyset\\
%\textsf{replace}(body_1, body_2)=body_1 & \textsf{if }body_1.\textsf{returnType}<:body_2.\textsf{returnType}\\
%\textsf{replace}(body_1, body_2)=\textsf{ERROR} \hspace{.1in} & \textsf{otherwise}
%\end{array}
%\end{equation*}

>>>>

\section{What  \mixin Generates}\label{sec:appendixtranslation}

This section presents a formal definition for most of the generated methods by $\mixinAnn$.

\subsection{Translation}\label{subsec:translation}

The translation functions of $\mixinAnn$ and $\weakAnn$
are presented in Figure~\ref{figure:translation}. Note that
it is necessary to explicitly check if the interface is valid
for annotation:

\noindent$\begin{array}{ll}
\InTextDef{7ex}{\valid(\C_0)}{\forall \m\in\dom(\C_0),\mbox{ if }\mh\QM; =
            \mBody(\m, \C_0),}\\
            \hspace{.24in}\mbox{ one of the following cases is satisfied:}\\
\tab\tab
\isField(\method), \isWith(\method, \C_0) \mbox{ or }
\isSetter(\method,\C_0)\\
\InTextDef{24ex}{\isField(\C\ \m\oR\cR\QM;)}{
\mnot\ \specialName(\m)}\\
\InTextDef{24ex}{\isWith(\C'\ \QM{with#}\m \oR \C\ \x\cR\QM;, \C_0)}{}\\
\hspace{.3in}\C_0 <: \C', \mBody(\m, \C_0) = \C\ \m\oR\cR\QM;
\ \mand\ \mnot\ \specialName(\m)\\
\InTextDef{24ex}{\isSetter(\C'\ \QM_\m \oR \C\ \x\cR\QM;, \C_0)}{}\\
\hspace{.3in}\C_0 <: \C', \mBody(\m, \C_0) = \C\ \m\oR\cR\QM;
\ \mand\ \mnot\ \specialName(\m)\\

@@

\noindent$\begin{array}{ll}
\InTextDef{16em}{\C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\QM;\in
\otherMethod(\C_0,\methods)}{}\\
\hspace{.25in}\isWith(\mBody(\QM{with#}\m, \C_0), \C_0),\ 
\QM{with#}\m\notin\dom(\methods)\\
\InTextDef{16em}{\C_0\ \QM_\m\oR \C\ \QM{_val}\cR\QM;\in
\otherMethod(\C_0,\methods)}{}\\
\hspace{.25in}\isSetter(\mBody(\QM_\m, \C_0), \C_0),\ \QM_\m\notin\dom(\methods)\\
\end{array}$

@@

\paragraph{The $\ofMethod$ function:}\label{subsec:ofmethod}
The function $\ofMethod$ generates the method \QM{of}, as an object factory. To
avoid boring digressions into well-known ways to find unique
names, %for the sake of this formalization
we assume that all methods with no parameters do not start with an underscore,
and we prefix method names with underscores to obtain valid parameter names for
\QM{of}.
\noindent$\begin{array}{l}
\InTextDef{5em}{\ofMethod(\C_0)}{
 \QM{static}\ \C_0\ \QM{of} \oR \C_1\ \QM_\m_1\QM,\ldots \C_n\ \QM_\m_n\cR\
\QM{\{}}\\
\tab\QM{return new}\ \C_0 \oR\cR\ \QM{\{}\\
\tab\tab \C_1\ \m_1 = \QM_\m_1\QM;\ldots \C_n\ \m_n = \QM_\m_n\QM; \\
\tab\tab
\C_1\ \m_1\oR\cR\ \QM{\{return }\ \m_1\QM{;\}}\ \ldots
\C_n\ \m_n\oR\cR\ \QM{\{return }\ \m_n\QM{;\}}\\
\tab\tab\withMethod(\C_1,\m_1,\C_0,\es_1)\ldots\withMethod(\C_n,\m_n,\C_0,\es_n)\\
\tab\tab\setterMethod(\C_1,\m_1,\C_0)\ldots\setterMethod(\C_n,\m_n,\C_0)\\
%\tab\tab\cloneMethod(\C_0,\es)\\
%\tab\tab\withMethod(\C_0)\\
\tab\QM{\};\}} \\
\InTextWith{\C_1\ \m_1\QM{();},\ldots \C_n\ \m_n\QM{();} = \fieldsFunc(\C_0)}\\
\hspace{.5in}\mbox{ and }\es_i=\m_1\QM,\ldots\QM, \m_{i-1}\QM,\QM{_val,}\m_{i+1}\QM,\ldots\QM, \m_n
\end{array}$

@@

\noindent$\begin{array}{ll}
\InTextDef{16ex}{\method\in\fieldsFunc(\C_0)}{}\\
\hspace{.3in}\isField(\method)\ \mand\
\method=\mBody(m^\method,\C_0)
\end{array}$

@@

\noindent$\begin{array}{l}
\InTextDef{10em}{\withMethod(\C,\m,\C_0,\es)}{}\\
\hspace{.2in}\C_0\ \QM{with#}\m\oR \C\ \QM{_val}\cR\ \QM{\{}
\QM{return}\ \C_0\QM{.of(}\es\QM{);\}} \\
\InTextWith{\mBody(\QM{with#}\m,\C_0) \mbox{ having the form }\mh\QM;}\\
\InTextDef{10em}{\withMethod(\C,\m,\C_0,\es)}{\emptyset\mbox{ otherwise}}\\
\InTextDef{10em}{\setterMethod(\C,\m,\C_0)}{}\\
\hspace{.2in}\C_0\ \QM_\m\oR \C\ \QM{_val}\cR\ \QM{\{}
 \m\QM{= _val;return this;\}} \\
\InTextWith{
\mBody(\QM_\m,\C_0) \mbox{ having the form }\mh\QM;}\\
\InTextDef{10em}{\setterMethod(\C,\m,\C_0)}{\emptyset\mbox{ otherwise}}\\
%\cloneMethod:&\cloneMethod(\C_0,\es)=
%\C_0\ \QM{clone()\{return}\ \C_0\QM{.of(}\es\QM{);\}} \\
%&\mbox{iff }
%\mBody(\QM{clone},\C_0) \mbox{ is of form }\mh\QM;\\
%&\cloneMethod(\C_0,\es)=\emptyset\mbox{ otherwise}\\
\end{array}$

\subsection{Other Features}\label{subsec:otherfeatures}

@@

### 2016-06-09 GPCE2016/sections/FutureWork.tex
preparing for GPCE

\noindent This is a safe and explicit way to list all the public interfaces and to hide all the implementation
details from outside of the package.\footnote{
Note how \Q@ExposedData.of@ does not need to map directly to the fields, and can also perform
invariant checking.}
Using packages as a boundary we can make all object interfaces \emph{where some form of privateness is desirable}
as invisible types for the rest of the program.
This preserves encapsulation, but limits inheritance:
package protected interfaces cannot be seen,
thus cannot be implemented from outside the package.
This can raise questions: keeping fields private is a staple of CB.
By forcing the user/heir to rely only on methods, we avoid (a part)
of the fragile base class problem and we obtain representation independence.
In IB fields do not exists; abstract state methods like getters and
setters are abstract: it does not make sense to make them private;
however, since they are methods, the base interface may be able to
evolve freely.  To understand if a fragile base interface problem
exists in IB, we need more research.

@@

\paragraph{A real IB language}
In this paper we presented the concept of IB as a programming pattern over Java.
An alternative would be to have a language providing IB as its only paradigm.
We do not wish to define here a complete IB language, just to show the reader IB as
a promising way to design a new OO language.
To define such language, the first task is to define the set of abstract state operations.
Of course we need getters; \Q@withX@ methods are also convenient, and they enable a
clean functional programming style that would be otherwise too cumbersome.
Currently we are offering both setters and fluent setters.
We believe fluent setters to be sufficient; we could drop the concept of (non fluent)setters.
Manually defining abstract state methods can be verbose, we could offer a convenient syntax to
declare in a compact way get/set/with.
For example:
\begin{lstlisting}
obj interface Foo{
  String bar;// expanded as getter and with:
  //String bar(); Foo withBar(String val);
  var String beer;//as before plus fluent setter
}
\end{lstlisting}

>>>>

\noindent This is a safe and explicit way to list all the public interfaces and to hide all the implementation
details from outside of the package.\footnote{
Note how \Q@ExposedData.of@ does not need to map directly to the fields, and can also perform
invariant checking.}
Using packages as a boundary we can make all object interfaces \emph{where some form of privateness is desirable}
as invisible types for the rest of the program.
This preserves encapsulation, but limits inheritance:
package protected interfaces cannot be seen,
thus cannot be implemented from outside the package.
This can raise questions: keeping fields private is a staple of CB.
By forcing the user/heir to rely only on methods, we avoid (a part)
of the fragile base class problem and we obtain representation independence.
In IB fields do not exist; abstract state methods like getters and
setters are abstract: it does not make sense to make them private;
however, since they are methods, the base interface may be able to
evolve freely.  To understand if a fragile base interface problem
exists in IB, we need more research.

@@

\paragraph{A real IB language}
In this paper we presented the concept of IB as a programming pattern over Java.
An alternative would be to have a language providing IB as its only paradigm.
We do not wish to define here a complete IB language, just to show the reader IB as
a promising way to design a new OO language.
To define such language, the first task is to define the set of abstract state operations.
Of course we need getters; \Q@withX@ methods are also convenient, and they enable a
clean functional programming style that would be otherwise too cumbersome.
Currently we are offering both setters and fluent setters.
We believe fluent setters to be sufficient; we could drop the concept of (non-fluent) setters.
Manually defining abstract state methods can be verbose, we could offer a convenient syntax to
declare in a compact way get/set/with.
For example:
\begin{lstlisting}
obj interface Foo{
  String bar;// expanded as getter and with:
  //String bar(); Foo withBar(String val);
  var String beer;//as before plus fluent setter
}
\end{lstlisting}
